Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-04-13 VegOzone.tex
Resturcture and edit Section 2

Ozone acts as oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, leaf pores that enable gas exchange, allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to key environmental variables such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself \citep[e.g.,][]{PTRS:Jarvis1976, BallBerry1987, Emberson2000, ICP:MappingManual2017}. Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Though, Scots pine (\emph{Pinus sylvestris}) is considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} found chlorophyll:carotenoid ratio and polyamines reductions under elevated ozone concentrations indicating susceptibility to ozone also in these species.\\

@@

%%
\section{Methods}
\label{sec:dam}
In the following, we briefly describe the model used for assessing the ozone risk, its key components, input data, and gap filling procedure (Section~\ref{sec:do3se}). To study the impact of model parameters, we propose a method for parameter localization to a subarctic climate based on existing Boreal parameterizations and estimate the start and end of the growing season from satellite derived GPP (Section~\ref{sec:localization}).

\subsection{DO3SE model}
\label{sec:do3se}
To investigate the \chem{O_3} damage risk in the subarctic, we model the ozone dose of natural (coniferous and deciduous trees) and semi-natural (grassland) vegetation using the Deposition of Ozone for Stomatal Exchange ($\mathrm{DO_3SE}$) model \citep{ACP:Bueker2012} which computes estimates of the total and stomatal deposition of \chem{O_3} to selected land-cover types and plant species according to UNECE LRTAP (Long-Range Transboundary Air Pollution) methodologies for effects-based risk assessment.

\subsubsection{Model description}
\label{subsec:model_describtion}
The opening of stomata is governed by different species-specific factors and response to key environmental variables. In $\mathrm{DO_3SE}$ it is described by a multiplicative empirical model \citep{PTRS:Jarvis1976,Emberson2000,ICP:MappingManual2017}:
\begin{equation}
  g^k_\text{sto} = g^k_\text{max} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_\mathrm{T} \cdot f^k_\mathrm{VPD} \cdot f^k_\text{SWP}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
Where $g_\mathrm{max}$ is a species-specific maximum stomatal conductance value that is then modified by seasonal and response functions to key environmental variables that vary within a range $0-1$. These factors are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_\mathrm{T}$), water vapor pressure deficit ($f_\mathrm{VPD}$), and soil water potential ($f_\text{SWP}$). All factors differ with plant functional type (PFT) denoted with $k$.

The phenology ($f_\mathrm{phen}$) is described by a stepwise defined linear function that allows for a increase of stomatal conductance in spring (start of growing season) and a decrease in fall (end of growing season). Furthermore, mid-season dips can also be accounted for.

The temperature adjustment function is defined as
%
\begin{equation}
  f_\mathrm{T} = \frac{T_\text{air}-T_\text{min}}{T_\text{opt}-T_\text{min}} \cdot \left(\frac{T_{\text{max}}-T_\text{air}}{T_{\text{max}}-T_\text{opt}}\right)^\beta,
  \label{eq:f_temp}
\end{equation}
with $\beta = \frac{T_\text{max}-T_\text{opt}}{T_\text{opt}-T_\text{min}}$ a mixed reciprocal polynomial function of temperature. The shape parameters $T_\text{min}$, $T_\text{max}$ and $T_\text{opt}$ are tabulated for various PFT and $T_\mathrm{air}$ is the near-surface air temperature. All temperatures are defined in units of \unit{^\circ C}.

The water vapor pressure deficit function is
%
\begin{equation}
  f_\mathrm{VPD} = f_\text{min}+(1-f_\text{min}) \cdot \frac{D_\text{min} - \text{VPD}}{D_\text{min}-D_\text{max}}
  \label{eq:f_vpd}
\end{equation}
%
where VPD is the leaf to air vapor pressure deficit in \unit{kPa} with $f_\text{min}$, $D_\text{min}$, $D_\text{max}$ describing the relative stomatal conductance to changes in vapor pressure deficit.

The wavelength band ($400-700\,\unit{nm}$) plant chlorophyll responds to is called photosynthetic active radiation (PAR). Its integral is the photosynthetic photon flux density (PPFD). The relationship between relative $g_\mathrm{sto}$ and PPFD is given by
%
\begin{equation}
  f_\text{light} = 1-\text{exp}(-\alpha_\text{light}\cdot \text{PPFD})
  \label{eq:flight}
\end{equation}
%
where $\alpha_\mathrm{light}$ is a slope parameter describing the extent of stomatal opening at low light intensities.

The $\mathrm{DO_3SE}$ model as described in \citet{ACP:Bueker2012} is used to simulate SWP across a PFT specific root depth according to the Penman--Monteith energy balance method that drives water cycling through the soil--plant--atmosphere system. The stomatal response to changes in SWP ($f_\mathrm{SWP}$) are, thus, given by:
%
\begin{equation}
  f_\mathrm{SWP} = \text{min}\left\{1, \text{max}\left\{f_\mathrm{min}, \frac{(1 - f_\mathrm{min})\cdot(\mathrm{SWP_{min}} - \mathrm{SWP})}{\mathrm{SWP_{min}} - \mathrm{SWP_{max}}} + f_\mathrm{min} \right\} \right\}
    \label{eq:fsw}
\end{equation}
%
where SWP is the soil water potential across the root zone and $\mathrm{SWP_{min}}$ and $\mathrm{SWP_{max}}$ are the parameters describing the $f_\mathrm{SWP}$ relationship.

$\Phi_\mathrm{sto}$ in Eq.~(\ref{eq:pod}) as computed by the $\mathrm{DO_3SE}$ model is a fraction of the total \chem{O_3} deposition flux. A standard resistance analogous approach \citep{AE:Wesely1989} is used to transport \chem{O_3} from the atmospheric reference height to the vegetated surface. The calculation of aerodynamic $R_a$ and quasi-laminar boundary layer resistance $R_b$ follow \citet{WASP:Simpson2003,ACP:Simpson2012}. The surface resistance $R_\mathrm{surf}$ consists of stomatal $r_\mathrm{sto}$ and non-stomatal canopy resistances that includes external plant surfaces $r_\mathrm{ext}$, aerodynamic within-canopy $R_\mathrm{inc}$ and ground surfaces or soil resistances $R_\mathrm{gs}$ \citep{WASP:Simpson2003,ACP:Simpson2012}. Under the assumption that \chem{[O_3]} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf, $r_\mathrm{sto}$ and $r_\mathrm{ext}$ are calculated per leaf/needle area (note lower case $r$).
%
\begin{equation}
  R_\mathrm{sur} = \frac{1}{\frac{\mathrm{LAI}}{r_\mathrm{sto}}+\frac{\mathrm{SAI}}{r_\mathrm{ext}}+\frac{1}{R_\mathrm{inc}+R_\mathrm{gs}}}
\end{equation}
%
Stomatal resistance is the inverse of stomatal conductance $g_\mathrm{sto}$ in Eq.~(\ref{eq:stomatal}). Accounting for these the effective stomatal flux $\Phi_\mathrm{sto}$ at the vegetation height $h$ is defined as
%
\begin{equation}
  \Phi_\mathrm{sto} = \chem{[O_3]}(h) \cdot g_\mathrm{sto}(h).
  \label{eq:flux_stomata}
\end{equation}
A linear gradient of \chem{[O_3]} towards the surface has been adopted in the $\mathrm{DO_3SE}$ model based on detailed studies using the chemical transport model EMEP \citep{ACP:Simpson2012,ICP:MappingManual2017}.

\subsubsection{Gap filling methodology}
\label{subsec:gap_filling}
The $\mathrm{DO_3SE}$ model requires hourly, continuous input data. Therfore, the following gap filling methodology was devised for the meteorological observations:
\begin{itemize}
\item Single hours of missing data were filled by taking the average of the hourly values before, and after, the missing value.
\item Several consecutive hours of missing data (23 or less) were filled by taking the average of the corresponding hour the day before, and the day after; and repeating this for each missing hour of data. If data were unavailable from that hour of the previous day, then only the value from the day after was used and vice versa. Interpolated values were not used in calculating averages.
\item Data gaps longer than $24\,\unit{h}$ were filled using weekly diurnal averages. i.e. an average was calculated using the corresponding hour throughout the week before and after. Interpolated values were not used in calculating averages.
\end{itemize}

To fill gaps in the ozone data, we applied a methodology based on a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with an ozone climatology \chem{\left<[O_3]\right>} and ozone anomalies (deviation from the norm) \chem{\Delta[O_3]}. We computed a generalized climatology for northern Fennoscandia from long-term records of different close by sites and for our site. We determined the station with the highest correlation coefficient compared to Svanhovd. For this site, we computed the anomalies with respect to the generalized climatology for the gap period. These anomalies were than added to the climatology at out site and corrected for an increase in present-day tropospheric background ozone compared to the early 1990s. For more details refer to \citet{ACP:Falk2021}.

\subsubsection{Input data}
\label{subsec:data}
The required input data consists of observed ozone concentration and meteorological observations of $2\,\unit{m}$ temperature $T_\mathrm{2m}$, global irradiation $Q_0$, precipitation, $2\,\unit{m}$ wind $u_\text{2m}$, and vapor pressure deficit VPD. VPD has been calculated from observed $T_\text{2m}$ and relative humidity
\begin{equation}
  \text{VPD} = P_s(T) \cdot \left(1-\frac{\text{relHum}}{100}\right),
\end{equation}
with saturation vapor pressure $P_s(T)$ in \unit{hPa} derived from Arden Buck equation~\citep{JAP:Buck1981, Buck2012}.

As site for our study, we chose Svanhovd ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$). Reasons for our choice: (1) Ambient ozone concentration had been monitored there in the past, (2) an agrometeorological station is measuring, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths, and (3) a national surveillance of air pollution (\chem{NO_x}, \chem{SO_x}, PM, radiological species) is ongoing. Most of these measurements are performed by the Norwegian Institute for Air Research (NILU), who also conducted ozone measurements on our behalf during the 2018/19 growing season. In addition, we collected atmospheric monitoring and ozone monitoring data taken at Svanhovd over the past $35\,\unit{years}$.

All measured essential climate variables for the 2018/19 growing season were provided by NILU through the web portal \href{luftkvalitet.no}{luftkvalitet.no} (last accessed May 2020). The ozone monitoring data have been added to a database for air-quality monitoring called EBAS as well operated by NILU and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Because ozone monitors did not undergo regular re-calibration drifts in the observed data occurred imposing false trends. This does not concern the general ozone seasonality which we are particularly interested in.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at Svanhovd. We report \chem{[O_3]} as volume mixing ratios (VMR) in \unit{ppb} for easier comparison with exposure-based risk assessment. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3).

@@

%%%

\subsection{Model parameter localization}
\label{sec:localization}

We deduced the dominant species on-site at Svanhovd from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Generic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations in the following. Our parameterization of coniferous trees is based on the MM's Boreal Norway spruce, deciduous trees on the MM's silver birch, and perennial grassland on MM's perennial grassland for central Europe. For a comprehensive list of generic MM model parameters, consult Supplement~Table~S1.

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance in this species occurred only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction that is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance in perennial grassland. Therefore, we saw the necessity to propose bespoke parameterizations describing an acclimation of the generic PFTs to the climatic conditions in the focus area. Note, however, that these parameterizations are hypothetical and have yet to be verified by experiments.

We presume that perennial vegetation will likely maximize carbon acquisition during the short subarctic growing season. Because the MM version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ to acclimate the growth potential to the local climate. For details regarding these functions see Appendix~\ref{appendix:do3se_model}. Further, we presume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suit our vegetation types and no acclimation is necessary for these.

We define two different hypothetical temperature acclimations: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. This is accomplished by moving $T_\mathrm{opt}$ towards cooler temperatures while keeping the other parameters fixed to MM values. In the same way, \emph{subarctic} is constructed to represent a species that is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift $T\mathrm{min}$ and $T\mathrm{max}$ to colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. In other words, we aim to increase the overlapping area between $f_\mathrm{T}$ and the temperature probability density function (PDF) during the GS. The most extreme of the two acclimation types is \emph{subarctic}.
In Fig.~\ref{fig:f_temp_grassland}a, we show the resulting $f_\mathrm{T}$ for perennial grassland together with temperature PDFs deduced from the five months GS in two climatological periods. The periods 1990s and 2000s have been chosen to indicate that the vegetation has been subject to climate change. Thus, the predominant acclimation at present is not known.

The resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}a and \ref{fig:f_temp_birch}a, respectively. We base our temperature acclimation of coniferous trees on experimental results on Norway spruce which were found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. In particular, we use the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007} to estimate the optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). We presume a similar temperature acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.\\

>>>>

Ozone acts as oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, leaf pores that enable gas exchange, allowing the entry of \chem{CO_2} for photosynthesis and loss of \chem{H_2O} vapor via the plant transpiration stream. Stomatal aperture, and hence stomatal conductance to these gases (including ozone) will vary over the day and growing season primarily to balance \chem{CO_2} uptake against \chem{H_2O} vapor loss. The higher the stomatal conductance, the higher the potential for ozone uptake. Stomatal conductance has been empirically linked to key environmental variables such as air temperature $T$, photosynthetic photon flux density (PPFD), vapor pressure deficit (VPD), and soil water potential (SWP) as well as photosynthesis itself \citep[e.g.,][]{PTRS:Jarvis1976, BallBerry1987, Emberson2000, ICP:MappingManual2017}. Results from open-top chamber (OTC) experiments on downy birch (\emph{Betula pubescens}) and mountain birch (\emph{Betula pubescens toruosa}), native to subarctic regions, indicated reductions in both biomass, root:shoot ratio, and visible leaf damage under elevated ozone treatment ($\left<\chem{[O_3]}\right> = 36-54\,\unit{ppb}$) \citep{Amb:Manninen2009}. Though, Scots pine (\emph{Pinus sylvestris}) is considered to be more ozone tolerant due to an absence of visible injuries \citep{Amb:Girgzdiene2009}, \citet{Amb:Manninen2009} found chlorophyll:carotenoid ratio and polyamines reductions under elevated ozone concentrations indicating susceptibility to ozone also in these species.

@@

%%
\section{Methods}
\label{sec:dam}
In the following, we briefly describe the model used for assessing the ozone risk, its key components, input data, and gap filling procedure (Section~\ref{sec:do3se}). To study the impact of model parameters, we propose a method for parameter localization to a subarctic climate based on existing Boreal parameterizations and local climate conditions and estimate the start and end of the growing season from satellite derived GPP (Section~\ref{sec:localization}).

\subsection{DO3SE model}
\label{sec:do3se}
In our investigation of the ozone risk in the subarctic, we use the well established Deposition of Ozone for Stomatal Exchange ($\mathrm{DO_3SE}$) model \citep{ACP:Bueker2012} which computes estimates of the total and stomatal deposition of \chem{O_3} to selected land-cover types and plant species according to UNECE Long-Range Transboundary Air Pollution (LRTAP) methodologies for effects-based risk assessment.

\subsubsection{Model description}
\label{subsec:model_describtion}

In the $\mathrm{DO_3SE}$ model, the stomatal flux $\Phi_\mathrm{sto}$ in Eq.~(\ref{eq:pod}) is a fraction of the total \chem{O_3} deposition flux. A standard resistance analogous approach \citep{AE:Wesely1989} is used to transport \chem{O_3} from the atmospheric reference height $h_0$ to the vegetated surface. The calculation of aerodynamic $R_a$ and quasi-laminar boundary layer resistance $R_b$ follow \citet{WASP:Simpson2003,ACP:Simpson2012}. The surface resistance $R_\mathrm{surf}$ consists of stomatal $r_\mathrm{sto}$ and non-stomatal canopy resistances that include external plant surfaces $r_\mathrm{ext}$, aerodynamic within-canopy $R_\mathrm{inc}$ and ground surfaces or soil resistances $R_\mathrm{gs}$ \citep{WASP:Simpson2003,ACP:Simpson2012}. Under the assumption that the ozone concentration \chem{[O_3]} at the top of the canopy represents a reasonable estimate of the concentration at the upper surface of the laminar layer for a sunlit upper canopy leaf, $r_\mathrm{sto}$ and $r_\mathrm{ext}$ are calculated per leaf/needle area (note lower case $r$).
%
\begin{equation}
  R_\mathrm{sur} = \frac{1}{\frac{\mathrm{LAI}}{r_\mathrm{sto}}+\frac{\mathrm{SAI}}{r_\mathrm{ext}}+\frac{1}{R_\mathrm{inc}+R_\mathrm{gs}}}
\end{equation}
%
Stomatal resistance $r_\mathrm{sto}$ is the inverse of stomatal conductance $g_\mathrm{sto}$ in Eq.~(\ref{eq:stomatal}). Accounting for these the effective stomatal flux $\Phi_\mathrm{sto}$ at the vegetation height $h$ is defined as
%
\begin{equation}
  \Phi_\mathrm{sto} = \chem{[O_3]}(h) \cdot g_\mathrm{sto}(h).
  \label{eq:flux_stomata}
\end{equation}
A linear gradient reducing \chem{[O_3]} from $h_0$ to $h$ was adopted in the $\mathrm{DO_3SE}$ model based on detailed transport studies using the chemical transport model EMEP \citep{ACP:Simpson2012,ICP:MappingManual2017}.

The opening of stomata or stomatal conductance $g_\mathrm{sto}$ is governed by different species-specific factors and response to key environmental variables. In the $\mathrm{DO_3SE}$ model, it is described by a multiplicative empirical model \citep{PTRS:Jarvis1976,Emberson2000,ICP:MappingManual2017}:
\begin{equation}
  g_\text{sto} = g_\text{max} \cdot f_\text{phen} \cdot f_\text{light} \cdot \max{\left\{f_\text{min}, f_\mathrm{T} \cdot f_\mathrm{VPD} \cdot f_\text{SWP}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
Where $g_\mathrm{max}$ is a species-specific maximum stomatal conductance value that is modified by factors that vary within a range of $0-1$. These factors are empirically determined and account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_\mathrm{T}$), water vapor pressure deficit ($f_\mathrm{VPD}$), and soil water potential ($f_\text{SWP}$). All factors differ with plant functional type (PFT).

The leaf phenology factor ($f_\mathrm{phen}$) is a stepwise defined linear function that allows for an increase of stomatal conductance in spring (start of growing season) and a decrease in fall (end of growing season). Mid anthesis and mid season dips in stomata conductance can also be accounted for but are not relevant in the subarctic.

The temperature response function is defined as
%
\begin{equation}
  f_\mathrm{T} = \frac{T_\text{air}-T_\text{min}}{T_\text{opt}-T_\text{min}} \cdot \left(\frac{T_{\text{max}}-T_\text{air}}{T_{\text{max}}-T_\text{opt}}\right)^\beta,
  \label{eq:f_temp}
\end{equation}
with $\beta = \frac{T_\text{max}-T_\text{opt}}{T_\text{opt}-T_\text{min}}$ a mixed reciprocal polynomial function of temperature. The shape parameters $T_\text{min}$, $T_\text{max}$ and $T_\text{opt}$ are tabulated for various PFT and $T_\mathrm{air}$ is the near-surface air temperature. All temperatures are defined in units of \unit{^\circ C}.

The water vapor pressure deficit function is
%
\begin{equation}
  f_\mathrm{VPD} = f_\text{min}+(1-f_\text{min}) \cdot \frac{D_\text{min} - \text{VPD}}{D_\text{min}-D_\text{max}}
  \label{eq:f_vpd}
\end{equation}
%
where VPD is the leaf to air vapor pressure deficit in \unit{kPa} with a minimum conductance $f_\text{min}$. $D_\text{min}$, $D_\text{max}$ describe the relative stomatal conductance to changes in vapor pressure deficit.

The light response function $f_\mathrm{light}$ is given by
%
\begin{equation}
  f_\text{light} = 1-\text{exp}(-\alpha_\text{light}\cdot \text{PPFD})
  \label{eq:flight}
\end{equation}
%
where $\alpha_\mathrm{light}$ is a slope parameter describing the extent of stomatal opening at low light intensities. The wavelength band $400-700\,\unit{nm}$ to which plant chlorophyll responds to is called photosynthetic active radiation (PAR). Its integral is the photosynthetic photon flux density (PPFD).

In the $\mathrm{DO_3SE}$ model, Soil Water Potential (SWP) across a PFT specific root depth is simulated according to the Penman--Monteith energy balance method that drives water cycling through the soil--plant--atmosphere system \citep{ACP:Bueker2012}. The stomatal response to changes in SWP ($f_\mathrm{SWP}$) are, thus, given by:
%
\begin{equation}
  f_\mathrm{SWP} = \text{min}\left\{1, \text{max}\left\{f_\mathrm{min}, \frac{(1 - f_\mathrm{min})\cdot(\mathrm{SWP_{min}} - \mathrm{SWP})}{\mathrm{SWP_{min}} - \mathrm{SWP_{max}}} + f_\mathrm{min} \right\} \right\}
    \label{eq:fsw}
\end{equation}
%
where $\mathrm{SWP_{min}}$ and $\mathrm{SWP_{max}}$ are the parameters describing the $f_\mathrm{SWP}$ relationship.

\subsubsection{Gap filling methodology}
\label{subsec:gap_filling}
The $\mathrm{DO_3SE}$ model requires hourly, continuous input data. Therefor, the following gap filling methodology was devised for the meteorological drivers of the model:
\begin{itemize}
\item Single hours of missing data were filled by taking the average of the hourly values before, and after, the missing value.
\item Several consecutive hours of missing data (23 or less) were filled by taking the average of the corresponding hour the day before, and the day after; and repeating this for each missing hour of data. If data were unavailable from that hour of the previous day, then only the value from the day after was used and vice versa. Interpolated values were not used in calculating averages.
\item Data gaps longer than $24\,\unit{h}$ were filled using weekly diurnal averages. i.e. an average was calculated using the corresponding hour throughout the week before and after. Interpolated values were not used in calculating averages.
\end{itemize}

To fill gaps in the ozone data, we applied the methodology described in \citet{ACP:Falk2021} that is based on a Reynolds decomposition of the form:
\begin{equation}
  \chem{[O_3]} = \chem{\left<[O_3]\right>} + \chem{\Delta[O_3]},
  \label{eq:reynolds_decomp}
\end{equation}
with an ozone climatology (multiannual norm) \chem{\left<[O_3]\right>} and ozone anomalies (deviation from the norm) \chem{\Delta[O_3]}. We computed a generalized climatology for northern Fennoscandia from long-term records of different close by sites and for our site. We determined the station with the highest correlation coefficient compared to our site. For this site, we calculated the anomalies with respect to the generalized climatology for the gap period. These anomalies were than added to the climatology of our site and corrected for an increase in present-day tropospheric background ozone compared to the early 1990s. For more details refer to \citet{ACP:Falk2021}.

\subsubsection{Input data and observation site}
\label{subsec:data}
The required input data consists of ozone concentration $\chem{[O_3]}$, $2\,\unit{m}$ temperature $T_\mathrm{2m}$, global irradiation $Q_0$, precipitation, $2\,\unit{m}$ wind $u_\mathrm{2m}$, and vapor pressure deficit VPD. VPD has been calculated from observed $T_\mathrm{2m}$ and relative humidity
\begin{equation}
  \mathrm{VPD} = P_s(T) \cdot \left(1-\frac{\mathrm{relHum}}{100}\right),
\end{equation}
with saturation vapor pressure $P_s(T)$ in \unit{hPa} from Arden~Buck~equation~\citep{JAP:Buck1981, Buck2012}.

As site for our study, we chose Svanhovd in northern Norway ($69.45\,\unit{^\circ N}$, $30.03\,\unit{^\circ E}$, $30\,\unit{m\,a.s.l.}$). Reasons for our choice: (1) Ambient ozone concentration had been monitored there in the past, (2) an agrometeorological station is measuring, e.g. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths, and (3) a national surveillance of air pollution (\chem{NO_x}, \chem{SO_x}, particular matter (PM), radiological species) is ongoing. Most of these measurements are performed by the Norwegian Institute for Air Research (NILU), who also conducted ozone measurements on our behalf during the 2018/19 growing season.

All essential climate variables for the 2018/19 growing season were provided by NILU through the web portal \href{luftkvalitet.no}{luftkvalitet.no} (last accessed May 2020). The ozone monitoring data have been added to a database for air-quality monitoring called EBAS as well operated by NILU and are thus openly accessible. Long-term ozone observation data (1986--1996) were obtained from EBAS. Note that data from before the 1990s do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Because ozone monitors did not undergo regular re-calibration drifts in the observed data occurred imposing false trends which are small but in the same order of magnitude as the expected trends in tropospheric background ozone. This does not concern the ozone seasonality which we are particularly interested in.
The long-term data sets of agrometeorological variables including temperature and precipitation are available from September 1992 to the present day \citep[LandbruksMeteorologiske Tjeneste][note the station name here is Pasvik]{LMT_NIBIO}.

To relate local ozone concentration \chem{[O_3]} with visible damage on ozone sensitive plants, an ozone bio-monitor (referred to as ozone garden) has been installed at Svanhovd in the 2018/19 GS. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clovers (\emph{Trifolium repens}, \emph{Trifolium pratense}), tobacco (\emph{Nicotiana tabacum}, cultivars Bel-W3, Bel-B and Bel-C), and potato (\emph{Solanum tuberosum}). As shown in Fig.~\ref{fig:svanhovd_research_station}b, the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar (Bel-W3).

@@

%%%
\subsection{Model parameter localization}
\label{sec:localization}

We deduced the dominant species on-site at Svanhovd from Fig.~\ref{fig:svanhovd_research_station}a and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Basic parameters for these PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as mapping manual (MM) parameterizations in the following. For coniferous trees, we base our localization of parameters on MM's Boreal Norway spruce. Our deciduous tree paramterization is based on MM's silver birch and perennial grassland on MM's perennial grassland for Central Europe. For a comprehensive list of generic MM model parameters, consult Supplement~Table~S1.

\subsubsection{Temperature acclimation}

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance in this species occurred only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction that is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low throughout the season, being the limiting factor of stomatal conductance in perennial grassland. Note, these parameterizations are hypothetical and have yet to be verified by experiments.

We presume that perennial vegetation will likely maximize carbon acquisition during the short subarctic growing season. Because the MM version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ to acclimate the growth potential to the local climate. For details regarding these functions see Section~\ref{subsec:model_describtion}. Further, we presume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suit our vegetation types and no acclimation is necessary for these. But acknowledge that these are very important factors as shown by~\citet{ACP:Bueker2012}.

We define two different hypothetical temperature acclimations: subarctic and cold. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. This is accomplished by moving $f_\mathrm{T}$ towards cooler temperatures but keep $T_\mathrm{max}$ at its MM value. In the same way, \emph{subarctic} is constructed to represent a species that is very tolerant to cold but sensitive to high temperatures, and most efficient at cool temperatures. Therefore, we shift $f\mathrm{T}$ to even colder temperatures and choose $T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000. In other words, we aim to increase the overlapping area between $f_\mathrm{T}$ and the temperature probability density function (PDF) during the GS. The most extreme of the two acclimation types is \emph{subarctic}.
In Fig.~\ref{fig:f_temp_grassland}a, we show the resulting $f_\mathrm{T}$ for perennial grassland together with temperature PDFs deduced from the five months GS in two climatological periods. The periods 1990s and 2000s have been chosen to indicate that the vegetation has been subject to climate change already. Thus, the predominant acclimation at present is not known.

The resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}a and \ref{fig:f_temp_birch}a, respectively. We base our temperature acclimation of coniferous trees on experimental results on Norway spruce which were found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. In particular, we use the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland by \citet{TB:Kolari2007} to estimate the optimal temperature interval ($10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$). We presume a similar temperature acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.
